Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems

### 2018-03-28 LaTeX/main_revised.tex
refining conclusions and reviewer 2 responses

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

@@

%reference equations in appendices
\usepackage{xr}
\externaldocument{S1Appendix}
\externaldocument{S2Appendix}
\externaldocument{S3Appendix}

@@

%Previous models with haploid selection
Two previous numerical simulation studies have considered the spread of GSD with sex-specific meiotic drive~\cite{Kozielska:2010vm,Ubeda:2015fx} under a small number of possible genetic architectures and diploid selective regimes. 
Ubeda et al. (2015)~\cite{Ubeda:2015fx}, considers ancestral-ESD (with no sex ratio bias) and find that new GSD alleles can spread if they arise in linkage with meiotic drive loci. 
For example, a masculinizing allele spreads in association with an allele that is favoured during male meiosis, causing sex ratios to become male-biased and thereby suggesting that such beneficial associations overwhelm selection to balance the sex ratio. 
However, Kozielska et al. (2010)~\cite{Kozielska:2010vm} consider meiotic drive on an ancestral GSD sex chromosome (creating a sex ratio bias). 
They find that new, unlinked, GSD systems can spread if they generate the rarer sex, creating a balanced sex ratio. 
That is, they conclude that Fisherian sex ratio selection caused by haploid selection favours new GSD systems. 
Here, we analytically find conditions under which new GSD or ESD systems spread with generic linkage between the loci involved (and generic sex-specific haploid and diploid selection).

%main results
Strikingly, we reconcile the results of Kozielska et al. (2010)~\cite{Kozielska:2010vm} and Ubeda et al. (2015)~\cite{Ubeda:2015fx} by showing that new GSD systems commonly spread when they increase \textit{or} decrease linkage with loci that experience haploid selection.
This result is qualitatively distinct from those for diploid selection alone~\cite{vanDoorn:2007eu,vanDoorn:2010hu} and suggests that haploid selection is more likely to promote transitions between sex-determination systems. 
However, we also show that transitions involving haploid selection cannot be simply explained by invoking sex ratio selection. 
In particular, under a wide range of conditions, we show that haploid selection drives transitions in sex-determining system \textit{equally strongly} whether sex-ratio biases increase or decrease (and whether sex-ratio biases are present or absent). 
Finally, we show that ESD may not evolve, even if the sex-ratio is initially biased by haploid selection, which would not be predicted by previous theories for transitions to ESD~\cite{Charnov:1982wg,Bull:1983vi,West:2009we}.

@@

%life-cycle
A full description of our model, including recursion equations, is given in \nameref{app:recurs}. 
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before gametic competition. 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition) separately. 
Selection during gametic competition depends on the $\mathbf{A}$ locus genotype; relative fitnesses are given by $w_A^\circ$ and $w_a^\circ$ ($\circ \in \{\female,\male\}$; see table \ref{tab:fitnesstable}). %$w_{A}^\male$ and $w_{a}^\male$ for male gametes and $w_{A}^\female$ and $w_{a}^\female$ for female gametes, see table \ref{tab:fitnesstable}. 
We assume that all gametes compete for fertilization during gametic competition, which implies a polygamous mating system. 
Gametic competition in monogamous mating systems is, however, equivalent to meiotic drive in our model (described below), as either only alters the frequency of gametes produced by heterozygotes. 
After gametic competition, conjugation between male and female gametes occurs at random.
The resulting zygotes develop as males or females, depending on their genotypes at the $\mathbf{X}$ and $\mathbf{M}$ loci. %(and the $\mathbf{M}$ genotype of their mother in the case of maternal control).
Diploid males and females then experience \textcolor{red}{viability(?)} selection, with relative fitnesses $w_{AA}^{\circ}$, $w_{Aa}^{\circ}$, and $w_{aa}^{\circ}$. %where $g$ is the diploid genotype at the $\mathbf{A}$ locus ($g \in \{AA, Aa, aa\}$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the $\mathbf{A}$ locus.
\textcolor{red}{We do not consider fertility(?) or sexual selection.}
The next generation of gametes is produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination (i.e., an odd number of cross-overs) occurs between loci $\mathbf{X}$ and $\mathbf{A}$ with probability $r$, between loci $\mathbf{A}$ and $\mathbf{M}$ with probability $R$, and between loci $\mathbf{X}$ and $\mathbf{M}$ with probability $\rho$.
Any linear order of the loci can be modelled with appropriate choices of $r$, $R$, and $\rho$ (see Fig \ref{fig:model_outline}A and \nameref{tab:chisubstitutions}). 
Individuals that are heterozygous at the $\mathbf{A}$ locus may experience meiotic drive; a gamete produced by $Aa$ heterozgotes of sex $\circ$ bears allele $A$ with probability $\alpha^\circ$. 
Thus, the $\mathbf{A}$ locus can experience sex-specific gametic competition, diploid selection, and/or meiotic drive. 
`
\begin{table}[ht]
%\centering
\smallskip
\caption{{\bf Relative fitness of different genotypes in sex $\circ \in \{\female,\male\}$ } }
\begin{tabular}{l c }
\hline\hline
  Genotype & Relative fitness during gametic competition \\ [0.5ex] \hline
  $A$ & $w_{A}^\circ = 1+t^\circ$ \\
  $a$ & $w_{a}^\circ = 1$ \\ [0.5ex] \hline
  Genotype & Relative fitness during diploid selection \\ [0.5ex] \hline
  $AA$ & $w_{AA}^\circ = 1+ s^\circ$ \\
  $Aa$ & $w_{Aa}^\circ = 1+h^\circ s^\circ$ \\
  $aa$ & $w_{aa}^\circ = 1$ \\ [0.5ex] \hline
  Genotype & Transmission during meiosis in $Aa$ heterozygotes \\ [0.5ex] \hline
  $A$ & $\alpha^\circ=1/2+\alpha_{\Delta}^{\circ}/2$ \\
  $a$ & $1-\alpha^\circ=1/2-\alpha_{\Delta}^{\circ}/2$ \\
  \hline \hline
  \label{tab:fitnesstable}
 \end{tabular}
\end{table}

@@

The evolution of a new sex-determining system requires that a rare mutant allele, $m$, at the novel sex-determining locus, $\mathbf{M}$, increases in frequency when rare. 
This is determined by the leading eigenvalue ($\lambda_m^{(XY)}$) of the system of eight equations describing the frequency of eggs and sperm carrying the $m$ allele in the next generation (equations \ref{eq:recursions}). %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
This system simplifies substantially for a dominant neo-Y ($k=0$) or neo-W ($k=1$), see \nameref{app:inv_cond}. 
The leading eigenvalue for a rare neo-Y or neo-W allele, $m\in\{Y',W'\}$, is the largest value of X that solves $x^2+ b x + c = 0$.  %$(\lambda_m^{(i)})^2+ b \lambda_m^{(i)} + c =0$. 
The coefficients are $b= - (\Lambda_{mA}^{(XY)} + \Lambda_{ma}^{(XY)})+(\chi_{mA}^{(XY)} + \chi_{ma}^{(XY)})$ and $c = (\Lambda_{mA}^{(XY)} - \chi_{mA}^{(XY)}) (\Lambda_{ma}^{(XY)} - \chi_{ma}^{(XY)}) - \chi_{mA}^{(XY)} \chi_{ma}^{(XY)}$, where $\Lambda_{mi}^{(XY)}>0$ is the multiplicative growth rate (which we will call the ``haplotypic growth rate'') of the neo-sex determination allele $m$ on background $i\in\{A,a\}$ without accounting for loss due to recombination ($R=0$), and $\chi_{mi}^{(XY)}>0$ is the rate at which mutant haplotypes on background $i\in\{A,a\}$ recombine onto the other $\mathbf{A}$ locus background in heterozygotes (proportional to $R$; \textcolor{red}{we call this the ``haplotypic recombination rate"(?)}), see Table \ref{tab:haplotype_growth}.
In the ancestral population, it is convenient to follow the frequency of the $A$ allele among female gametes (eggs), $p^\female_X$, and among X-bearing, $p^\male_X$, and among Y-bearing, $p^\male_Y$, male gametes (sperm/pollen). 
We also track the fraction of male gametes that are Y-bearing, $q$, which may deviate from $1/2$ due to meiotic drive in males. 
We will consider only equilibrium frequencies of alleles, $\hat{p}^\circ_i$, and Y-bearing male gametes, $\hat{q}$, when calculating the eigenvalues.

@@

The new sex-determining allele increases in frequency when rare when the largest eigenvalue is greater than one ($\lambda_m^{(XY)}>1$). 
%If the average change in frequency of the two haplotypes that carry the $m$ allele ($Am$ and $am$) is positive, invasion will always occur, i.e., if $(\Lambda_{mA} + \Lambda_{ma})/2 > 1$ then $\Lambda > 1$. 
%The leading eigenvalue is the largest solution of $f(\lambda_m^{(i)})=(\lambda_m^{(i)})^2+b\lambda_m^{(i)}+c=0$ and the Perron-Frobenius theorem guarantees that it is positive, unique, and real.  
%Since $f(\Lambda_{mA}^{(i)})$ and $f(\Lambda_{ma}^{(i)})$ are of opposite signs, the leading eigenvalue must fall between these two quantities and is the larger of them when $R = 0$ (see \nameref{file:Mathematica}).  
If both haplotypic growth rates are greater than one ($\Lambda_{mA}^{(XY)}, \Lambda_{ma}^{(XY)}>1$), then the new sex-determining allele invades regardless the rate of recombination between the new sex-determining locus and the selected locus ($R$), see \nameref{app:inv_cond} for details.
Conversely, if both haplotypic growth rates are less than one ($\Lambda_{mA}^{(XY)}, \Lambda_{ma}^{(XY)}<1$), then invasion can never occur.  
Finally, if only one haplotypic growth rate is greater than one, the new sex-determining allele can always invade when arising at a locus that is tightly linked to the selected locus ($R\approx0$).  
Furthermore, it can be shown that the leading eigenvalue declines with $R$, and invasion requires that $R$ is sufficiently small such that: 
\begin{equation}\label{eq:lambdasGen}
%\frac{\chi_{mA}}{\Lambda_{mA}-1}+\frac{\chi_{ma}}{\Lambda_{ma}-1}>0.
\chi_{ma}^{(XY)}/\left(\Lambda_{ma}^{(XY)} - 1\right) + \chi_{mA}^{(XY)}/\left(\Lambda_{mA}^{(XY)} - 1\right) < 1.
\end{equation}
\noindent 
This condition may or may not be satisfied for the full range of locations of the new sex-determining locus, including $R=1/2$, depending on the nature of selection.  
Interpreting this condition, if we assume that only the $mA$ haplotype would increase in frequency when $R=0$ ($\Lambda_{ma}^{(XY)}<1<\Lambda_{mA}^{(XY)}$) the first term on the left-hand side of \eqref{eq:lambdasGen} is negative and invasion requires that rate of growth of $mA$ haplotypes ($\Lambda_{mA}^{(i)}-1>0$) and the rate at which $mA$ haplotypes are produced by recombination in $ma$ haplotypes ($\chi_{ma}^{(i)}$) are sufficiently large relative to the rate of decline of $ma$ haplotypes ($1-\Lambda_{ma}^{(i)}>0$) and the rate of loss of $mA$ haplotypes due to recombination ($\chi_{mA}^{(i)}$).

We draw three main points about the generic invasion of neo-Y and neo-W mutations from Table \ref{tab:haplotype_growth}.
First, invasion by a neo-Y (neo-W) does not directly depend on the fitness of female (male) diploids. %(for a given set of equilibrium allele frequencies).
This is because a dominant neo-Y (neo-W) is always found in males (females), and therefore the frequency of the neo-Y (neo-W), $m$, only changes in males (females), Fig \ref{fig:model_outline}C,D.
Second, Fisherian sex-ratio selection will favour the spread of a neo-W and inhibit the spread of a neo-Y if the ancestral zygotic sex ratio is biased towards males (i.e., the first factor of the $\Lambda_{mi}^{(XY)}$ is greater than one for a neo-W and less than one for a neo-Y when $\zeta>1/2$), as might occur when the ancestral sex-determining locus is linked to a locus experiencing meiotic drive in males, for example.
However, the spread of a neo-Y (neo-W) also depends on the male (female) fitness of associated alleles (see terms involving equilibrium allele frequencies, $\hat{p}$'s). That is, 
\vspace{0.5cm}

\noindent\textit{Conclusion 1:} 
\textit{neither sex-ratio selection or selection on associated alleles dominates the invasion dynamics of a new sex-determining allele, allowing the sex ratio to become more or less biased during a transition.}
\textcolor{red}{[as suspected from Kozielska and Ubeda special cases]}
\vspace{0.5cm}

\noindent And thirdly, Table \ref{tab:haplotype_growth} also shows that
\vspace{0.5cm}

\noindent\textit{Conclusion 2: cis- and trans-GSD transitions are qualitatively different.}
\textcolor{red}{[already known from van Doorn and Kirkpatrick?]}
\vspace{0.5cm}

\noindent This is because, in an ancestrally XY system, a gamete with the neo-Y always pairs with a female gamete containing an X, Fig \ref{fig:model_outline}C.
By contrast, a gamete with a neo-W can pair with an X- or Y-bearing male gamete, Fig \ref{fig:model_outline}D.
Consequently, neo-W-bearing females obtain a different frequency of $A$ alleles from mating (when $\hat{p}_X^\male \neq \hat{p}_Y^\male$) compared to ancestral ($MM$) females, which can inhibit or favour its spread.

@@

\noindent\textit{Conclusion 3: selection on loci within the non-recombining region around the sex-determining locus ($r\approx0$) allows the invasion of less closely linked sex-determining alleles ($r<R\leq1/2$) during a trans-GSD transition}.
\textcolor{red}{[new result for sex-antagonism and overdominance, but already shown by Kozielska for ploidy-antagonism]}
\vspace{0.5cm}

\noindent This does not depend on the form of selection maintaining a polymorphism (sexually-antagonistic selection, overdominance, ploidally-antagonistic selection, or some combination, Fig \ref{fig:SexAntagTighter}, \nameref{fig:temporalOverdominance}, \nameref{fig:regionPloidAntag}, and \nameref{fig:SexAntagTighterMaleDrive}).  
The conditions become more restrictive, however, with increasing recombination ($R$) between the new sex-determining locus and the selected locus.  
The invasion of completely unlinked neo-W alleles ($R = 1/2$) can occur with overdominance in males or with haploid selection but is not possible with only sexually-antagonistic selection if selection is directional in each diploid sex (see \nameref{file:Mathematica}).
\vspace{0.5cm}

@@

\noindent\textit{Conclusion 4: when there is an intermediate rate of recombination between the ancestral sex-determining allele and a selected locus ($0<r<1/2$), haploid selection allows less closely linked sex-determining alleles ($r<R$) to invade during trans-GSD transitions ($\lambda_{W'}^{(XY)}>1$).}
\textcolor{red}{[new result because Kozielska had tight linkage and Ubeda had ESD]}
\vspace{0.5cm}

@@

Previous research has shown that when the ancestral sex-determining locus is linked to a locus that experiences haploid selection (e.g., meiotic drive), alleles at a new, unlinked sex-determining locus can invade and restore equal sex ratios~\cite{Kozielska:2010vm,Ubeda:2015fx}. 
Consider, for example, the case where the $\mathbf{A}$ locus is linked to the ancestral sex-determining locus ($r<1/2$) and experiences meiotic drive in males only ($\alpha^\male_\Delta \neq 0$, $\alpha^\female_\Delta=0$), without gametic competition ($t^\female=t^\male=0$).
In this case, the zygotic sex ratio can be initially biased only if the ancestral sex-determining system is XY (Fig \ref{fig:model_outline}A and Fig \ref{fig:SexRatioBad}B). 
%If the ancxestral sex-determining system is ZW, the zygotic sex ratio will be 1:1 because diploid sex is determined by the proportion of Z-bearing versus W-bearing eggs and meiosis in females is fair (Fig \ref{fig:Combination_Turnover}D).
If Fisherian sex-ratio selection were dominant, we would thus expect a difference in the potential for XY to ZW and ZW to XY transitions. 
%However, to leading order with selection weak relative to recombination, we find that sex-ratio selection favours the spread of a neo-W by an amount equl to the spread of a neo-Y into a ZW system (relabeling XY as the heterogametic sex and reversing the labels of males and females in Table \ref{tab:specialcases})
%However, to leading order with selection weak relative to recombination, we find that sex-ratio selection favours the spread of a neo-W (through the first terms in table \ref{tab:haplotype_growth}) by an amount that is equal in magnitude to the fitness effects of alleles associated with new sex-determining alleles (second terms in table \ref{tab:haplotype_growth}).
However, invasion by a neo-W allele into an XY system and invasion by a neo-Y allele into a ZW system occur under the same conditions ($\lambda_{Y'}^{(XY)} = \lambda_{W'}^{(ZW)}$ and $\lambda_{W'}^{(XY)}=\lambda_{Y'}^{(ZW)}$, at least to order $\epsilon^2$).
For example, in Fig \ref{fig:SexRatioBad}A neo-W alleles invade an ancestral-XY system where females are initially rare, equalizing the sex ratio.
However, Fig \ref{fig:SexRatioBad}B shows that a neo-Y can invade the resulting ZW system under the same conditions. 
When $R<1/2$, the invading neo-Y becomes associated with the male meiotic drive allele and the zygotic sex ratio actually evolves to become male-biased (as in the scenarios considered in \cite{Kozielska:2010vm,Ubeda:2015fx}).

>>>>

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath}%amssymb

@@

%reference equations in appendices
\usepackage{xr}
\externaldocument{S1Appendix_revised}
\externaldocument{S2Appendix_revised}
\externaldocument{S3Appendix_revised}

@@

%Previous models with haploid selection
Two previous studies have considered the spread of GSD with sex-specific meiotic drive~\cite{Kozielska:2010vm,Ubeda:2015fx} under a limited number of possible genetic architectures and diploid selective regimes. 
Ubeda et al. (2015)~\cite{Ubeda:2015fx}, considers ancestral-ESD (with no sex-ratio bias) and numerically show that new GSD alleles can spread if they arise in linkage with meiotic drive loci. 
For example, a masculinizing allele spreads in association with an allele that is favoured during male meiosis, causing sex ratios to become male-biased and thereby suggesting that such beneficial associations overwhelm selection to balance the sex ratio. 
However, Kozielska et al. (2010)~\cite{Kozielska:2010vm} consider meiotic drive on an ancestral GSD sex chromosome (creating a sex ratio bias). 
They find that new, completely unlinked, GSD systems can spread if they generate the rarer sex, creating a balanced sex ratio. 
That is, they conclude that Fisherian sex-ratio selection caused by haploid selection favours new GSD systems.

%main results
Here, we analytically find the conditions under which new GSD or ESD systems spread with generic linkage between the loci involved (and generic sex-specific haploid and diploid selection). 
Strikingly, we reconcile the results of Kozielska et al. (2010)~\cite{Kozielska:2010vm} and Ubeda et al. (2015)~\cite{Ubeda:2015fx} by showing that new GSD systems commonly spread when they increase \textit{or} decrease linkage with loci that experience haploid selection.
This result is qualitatively distinct from those for diploid selection alone~\cite{vanDoorn:2007eu,vanDoorn:2010hu} and suggests that haploid selection is more likely to promote transitions between sex-determination systems. 
However, we also show that transitions involving haploid selection cannot be simply explained by invoking sex-ratio selection. 
In particular, under a wide range of conditions, we show that haploid selection drives transitions in sex-determining system \textit{equally strongly} whether sex-ratio biases increase or decrease (and whether sex-ratio biases are present or absent). 
Finally, we show that ESD may not evolve, even if the sex ratio is initially biased by haploid selection, which would not be predicted by previous theories for transitions to ESD~\cite{Charnov:1982wg,Bull:1983vi,West:2009we}.

@@

%life-cycle
A full description of our model, including recursion equations, is given in \nameref{app:recurs}. 
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before gametic competition. 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition) separately. 
Selection during gametic competition depends on the $\mathbf{A}$ locus genotype; relative fitnesses are given by $w_A^\circ$ and $w_a^\circ$ ($\circ \in \{\female,\male\}$; see table \ref{tab:fitnesstable}). %$w_{A}^\male$ and $w_{a}^\male$ for male gametes and $w_{A}^\female$ and $w_{a}^\female$ for female gametes, see table \ref{tab:fitnesstable}. 
We assume that all gametes compete for fertilization during gametic competition, which implies a polygamous mating system. 
Gametic competition in monogamous mating systems is, however, equivalent to meiotic drive in our model (described below), as either only alters the frequency of gametes produced by heterozygotes. 
After gametic competition, conjugation between male and female gametes occurs at random.
The resulting zygotes develop as males or females, depending on their genotypes at the $\mathbf{X}$ and $\mathbf{M}$ loci. %(and the $\mathbf{M}$ genotype of their mother in the case of maternal control).
Diploid males and females then experience \textcolor{red}{viability(?)} selection, with relative fitnesses $w_{AA}^{\circ}$, $w_{Aa}^{\circ}$, and $w_{aa}^{\circ}$. %where $g$ is the diploid genotype at the $\mathbf{A}$ locus ($g \in \{AA, Aa, aa\}$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the $\mathbf{A}$ locus.
\textcolor{red}{We do not consider fertility(?) or sexual selection.}
The next generation of gametes is produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination (i.e., an odd number of cross-overs) occurs between loci $\mathbf{X}$ and $\mathbf{A}$ with probability $r$, between loci $\mathbf{A}$ and $\mathbf{M}$ with probability $R$, and between loci $\mathbf{X}$ and $\mathbf{M}$ with probability $\rho$.
Any linear order of the loci can be modelled with appropriate choices of $r$, $R$, and $\rho$ (see Fig \ref{fig:model_outline}A and \nameref{tab:chisubstitutions}). 
Individuals that are heterozygous at the $\mathbf{A}$ locus may experience meiotic drive; a gamete produced by $Aa$ heterozgotes of sex $\circ$ bears allele $A$ with probability $\alpha^\circ$. 
Thus, the $\mathbf{A}$ locus can experience sex-specific gametic competition, diploid selection, and/or meiotic drive.
The model is entirely deterministic and hence ignores chance fluctuations in allele frequencies due to genetic drift.

\begin{table}[ht]
%\centering
\smallskip
\caption{{\bf Relative fitness of different genotypes in sex $\circ \in \{\female,\male\}$ } }
\begin{tabular}{l c }
\hline\hline
  Genotype & Relative fitness during gametic competition \\ [0.5ex] \hline
  $A$ & $w_{A}^\circ = 1+t^\circ$ \\
  $a$ & $w_{a}^\circ = 1$ \\ [0.5ex] \hline
  Genotype & Relative fitness during diploid selection \\ [0.5ex] \hline
  $AA$ & $w_{AA}^\circ = 1+ s^\circ$ \\
  $Aa$ & $w_{Aa}^\circ = 1+h^\circ s^\circ$ \\
  $aa$ & $w_{aa}^\circ = 1$ \\ [0.5ex] \hline
  Genotype & Transmission during meiosis in $Aa$ heterozygotes \\ [0.5ex] \hline
  $A$ & $\alpha^\circ=1/2+\alpha_{\Delta}^{\circ}/2$ \\
  $a$ & $1-\alpha^\circ=1/2-\alpha_{\Delta}^{\circ}/2$ \\
  \hline \hline
  \label{tab:fitnesstable}
 \end{tabular}
\end{table}

@@

The evolution of a new sex-determining system requires that a rare mutant allele, $m$, at the novel sex-determining locus, $\mathbf{M}$, increases in frequency when rare. 
This is determined by the leading eigenvalue ($\lambda_m^{(XY)}$) of the system of eight equations describing the frequency of eggs and sperm carrying the $m$ allele in the next generation (equations \ref{eq:recursions}). %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
This system simplifies substantially for a dominant neo-Y ($k=0$) or neo-W ($k=1$), see \nameref{app:inv_cond}. 
The leading eigenvalue for a rare neo-Y or neo-W allele, $m\in\{Y',W'\}$, is the largest value of X that solves $x^2+ b x + c = 0$.  %$(\lambda_m^{(i)})^2+ b \lambda_m^{(i)} + c =0$. 
The coefficients are $b= - (\Lambda_{mA}^{(XY)} + \Lambda_{ma}^{(XY)})+(\chi_{mA}^{(XY)} + \chi_{ma}^{(XY)})$ and $c = (\Lambda_{mA}^{(XY)} - \chi_{mA}^{(XY)}) (\Lambda_{ma}^{(XY)} - \chi_{ma}^{(XY)}) - \chi_{mA}^{(XY)} \chi_{ma}^{(XY)}$, where $\Lambda_{mi}^{(XY)}>0$ is the multiplicative growth rate (which we will call the ``haplotypic growth rate'') of the neo-sex determination allele $m$ on background $i\in\{A,a\}$ without accounting for loss due to recombination ($R=0$), and $\chi_{mi}^{(XY)}>0$ is the rate at which mutant haplotypes on background $i\in\{A,a\}$ recombine onto the other $\mathbf{A}$ locus background in heterozygotes (proportional to $R$; we call this the ``dissociative force" as it breaks down linkage disequilibrium), see Table \ref{tab:haplotype_growth}.
In the ancestral population, it is convenient to follow the frequency of the $A$ allele among female gametes (eggs), $p^\female_X$, and among X-bearing, $p^\male_X$, and among Y-bearing, $p^\male_Y$, male gametes (sperm/pollen). 
We also track the fraction of male gametes that are Y-bearing, $q$, which may deviate from $1/2$ due to meiotic drive in males. 
We will consider only equilibrium frequencies of alleles, $\hat{p}^\circ_i$, and Y-bearing male gametes, $\hat{q}$, when calculating the eigenvalues.

@@

The new sex-determining allele increases in frequency when rare when the largest eigenvalue is greater than one ($\lambda_m^{(XY)}>1$). 
%If the average change in frequency of the two haplotypes that carry the $m$ allele ($Am$ and $am$) is positive, invasion will always occur, i.e., if $(\Lambda_{mA} + \Lambda_{ma})/2 > 1$ then $\Lambda > 1$. 
%The leading eigenvalue is the largest solution of $f(\lambda_m^{(i)})=(\lambda_m^{(i)})^2+b\lambda_m^{(i)}+c=0$ and the Perron-Frobenius theorem guarantees that it is positive, unique, and real.  
%Since $f(\Lambda_{mA}^{(i)})$ and $f(\Lambda_{ma}^{(i)})$ are of opposite signs, the leading eigenvalue must fall between these two quantities and is the larger of them when $R = 0$ (see \nameref{file:Mathematica}).  
If both haplotypic growth rates are greater than one ($\Lambda_{mA}^{(XY)}, \Lambda_{ma}^{(XY)}>1$), then the new sex-determining allele invades regardless the rate of recombination between the new sex-determining locus and the selected locus ($R$), see \nameref{app:inv_cond} for details.
Conversely, if both haplotypic growth rates are less than one ($\Lambda_{mA}^{(XY)}, \Lambda_{ma}^{(XY)}<1$), then invasion can never occur.  
Finally, if only one haplotypic growth rate is greater than one, the new sex-determining allele can always invade when arising at a locus that is tightly linked to the selected locus ($R\approx0$).  
Furthermore, it can be shown that the leading eigenvalue declines with $R$, and invasion requires that $R$ is sufficiently small such that: 
\begin{equation}\label{eq:lambdasGen}
%\frac{\chi_{mA}}{\Lambda_{mA}-1}+\frac{\chi_{ma}}{\Lambda_{ma}-1}>0.
\chi_{ma}^{(XY)}/\left(\Lambda_{ma}^{(XY)} - 1\right) + \chi_{mA}^{(XY)}/\left(\Lambda_{mA}^{(XY)} - 1\right) < 1.
\end{equation}
\noindent 
This condition may or may not be satisfied for the full range of locations of the new sex-determining locus, including $R=1/2$, depending on the nature of selection.  
Interpreting this condition, if we assume that only the $mA$ haplotype would increase in frequency when $R=0$ ($\Lambda_{ma}^{(XY)}<1<\Lambda_{mA}^{(XY)}$) then the first term on the left-hand side of \eqref{eq:lambdasGen} is negative and invasion requires that growth rate of $mA$ haplotypes ($\Lambda_{mA}^{(i)}-1>0$) and the rate at which they are produced by recombination ($\chi_{ma}^{(i)}$) are sufficiently large relative to the rate of decline of $ma$ haplotypes ($1-\Lambda_{ma}^{(i)}>0$) and the rate at which $m$ and $A$ are dissociated by recombination ($\chi_{mA}^{(i)}$).

We draw three main points about the generic invasion of neo-Y and neo-W mutations from Table \ref{tab:haplotype_growth}.
First, invasion by a neo-Y (neo-W) does not directly depend on the fitness of female (male) diploids. %(for a given set of equilibrium allele frequencies).
This is because a dominant neo-Y (neo-W) is always found in males (females), and therefore the frequency of the neo-Y (neo-W), $m$, only changes in males (females), Fig \ref{fig:model_outline}C,D.
Second, Fisherian sex-ratio selection will favour the spread of a neo-W and inhibit the spread of a neo-Y if the ancestral zygotic sex ratio is biased towards males (i.e., the first factor of the $\Lambda_{mi}^{(XY)}$ is greater than one for a neo-W and less than one for a neo-Y when $\zeta>1/2$), as might occur when the ancestral sex-determining locus is linked to a locus experiencing meiotic drive in males, for example.
However, the spread of a neo-Y (neo-W) also depends on the male (female) fitness of associated alleles (see terms involving equilibrium allele frequencies, $\hat{p}$'s).
This implies that neither sex-ratio selection or selection on associated alleles dominates the invasion dynamics of a new sex-determining allele, allowing the sex ratio to become more or less biased during a transition (as previously shown in two special cases; \cite{Kozielska:2010vm,Ubeda:2015fx}). 
%That is, 
%\vspace{0.5cm}
%
%\noindent\textit{Conclusion 1:} 
%\textbf{neither sex-ratio selection or selection on associated alleles dominates the invasion dynamics of a new sex-determining allele, allowing the sex ratio to become more or less biased during a transition.}
%\textcolor{red}{[as suspected from Kozielska and Ubeda special cases]}
%\vspace{0.5cm}
%
And thirdly, Table \ref{tab:haplotype_growth} also shows that
%\vspace{0.5cm}
%
%\noindent\textit{Conclusion 2:}
%\textbf{cis- and trans-GSD transitions are qualitatively different.}
%\textcolor{red}{[already known from van Doorn and Kirkpatrick]}
%\vspace{0.5cm}
%
cis- and trans-GSD transitions are qualitatively different.
This is because, in an ancestrally XY system, a gamete with the neo-Y always pairs with a female gamete containing an X, Fig \ref{fig:model_outline}C.
By contrast, a gamete with a neo-W can pair with an X- or Y-bearing male gamete, Fig \ref{fig:model_outline}D.
Consequently, neo-W-bearing females obtain a different frequency of $A$ alleles from mating (when $\hat{p}_X^\male \neq \hat{p}_Y^\male$) compared to ancestral ($MM$) females, which can inhibit or favour its spread.

@@

\noindent\textit{Conclusion 1A:}
\textbf{selection on loci within the non-recombining region around the sex-determining locus ($r\approx0$) allows the invasion of less closely linked sex-determining alleles ($r<R\leq1/2$) during a trans-GSD transition}.
%\textcolor{red}{[new result for sex-antagonism and overdominance, but already shown by Kozielska for ploidy-antagonism]}
\vspace{0.5cm}

\noindent This does not depend on the form of selection maintaining a polymorphism (sexually-antagonistic selection, overdominance, ploidally-antagonistic selection, or some combination, Fig \ref{fig:SexAntagTighter}, \nameref{fig:temporalOverdominance}, \nameref{fig:regionPloidAntag}, and \nameref{fig:SexAntagTighterMaleDrive}).  
The conditions become more restrictive, however, with increasing recombination ($R$) between the new sex-determining locus and the selected locus.  
The invasion of completely unlinked neo-W alleles ($R = 1/2$) can occur with overdominance in males or with haploid selection (as shown for the case of male drive opposed by diploid selection in \cite{Kozielska:2010vm}) but is not possible with only sexually-antagonistic selection if selection is directional in each diploid sex (see \nameref{file:Mathematica}).
%\vspace{0.5cm}

@@

\noindent\textit{Conclusion 1B:}\
\textbf{when there is an intermediate rate of recombination between the ancestral sex-determining allele and a selected locus ($0<r<1/2$), haploid selection allows less closely linked sex-determining alleles ($r<R$) to invade during trans-GSD transitions ($\lambda_{W'}^{(XY)}>1$).}
%\textcolor{red}{[new result because Kozielska had ancestrally-tight sex-linkage and Ubeda had ESD]}
\vspace{0.5cm}

@@

Previous research has shown that when the ancestral sex-determining locus is linked to a locus that experiences haploid selection (e.g., meiotic drive), alleles at a new, unlinked sex-determining locus can invade and restore equal sex ratios~\cite{Kozielska:2010vm}. 
Consider, for example, the case where the $\mathbf{A}$ locus is linked to the ancestral sex-determining locus ($r<1/2$) and experiences meiotic drive in males only ($\alpha^\male_\Delta \neq 0$, $\alpha^\female_\Delta=0$), without gametic competition ($t^\female=t^\male=0$).
In this case, the zygotic sex ratio can be initially biased only if the ancestral sex-determining system is XY (Fig \ref{fig:model_outline}A and Fig \ref{fig:SexRatioBad}B). 
%If the ancxestral sex-determining system is ZW, the zygotic sex ratio will be 1:1 because diploid sex is determined by the proportion of Z-bearing versus W-bearing eggs and meiosis in females is fair (Fig \ref{fig:Combination_Turnover}D).
If Fisherian sex-ratio selection were dominant, we would thus expect a difference in the potential for XY to ZW and ZW to XY transitions. 
%However, to leading order with selection weak relative to recombination, we find that sex-ratio selection favours the spread of a neo-W by an amount equl to the spread of a neo-Y into a ZW system (relabeling XY as the heterogametic sex and reversing the labels of males and females in Table \ref{tab:specialcases})
However, invasion by a neo-W allele into an XY system and invasion by a neo-Y allele into a ZW system occur under the same conditions ($\lambda_{Y'}^{(XY)} = \lambda_{W'}^{(ZW)}$ and $\lambda_{W'}^{(XY)}=\lambda_{Y'}^{(ZW)}$, at least to order $\epsilon^2$), implying that
\vspace{0.5cm}

\noindent\textit{Conclusion 2:}
\textbf{when selection is weak relative to recombination, sex-ratio selection is exactly equal in magnitude to the selection imposed by associations with selected alleles.}
%\textcolor{red}{[new result]}
\vspace{0.5cm}

\noindent For example, in Fig \ref{fig:SexRatioBad}A neo-W alleles invade an ancestral-XY system where females are initially rare, equalizing the sex ratio.
However, Fig \ref{fig:SexRatioBad}B shows that a neo-Y can invade the resulting ZW system under the same conditions. 
When $R<1/2$, the invading neo-Y becomes associated with the male meiotic drive allele and the zygotic sex ratio actually evolves to become male-biased (as in the scenarios considered in \cite{Kozielska:2010vm,Ubeda:2015fx}).
In this case, the neo-Y spreads because it is often found in males and can, if it carries the driven allele $a$, benefit from haploid selection in males (Fig \ref{fig:SexRatioBad}B).
